---
title: "WUP_04"
author: "Shawn Olson"
date: "fall 2024"
output:
  html_document:
    df_print: paged
---

```{r include=FALSE,echo=FALSE}
require(knitr)
require(tidyverse)
require(tigerstats)
GSScsv <- read_csv("GSScsv.csv")

```

# Introduction

<!--In this section you explain what you are trying to show.  Where did the data come from?  What is the research or other question you are trying to answer?!-->

The data is in the file GSScsv.csv and came from the 2018 General Social Survey.  The question to answer is to resolve this hypothesis.  
$H_0$ There is no difference between the average score on the WORDSUM test and level of schooling DEGREE  
$H_a$ There is a significant difference in average score on the WORDSUM test between at least two levels of DEGREE  

# Methods

<!--Mention that you intend to use chatGPT to the greatest extent possible. Give credit to chatGPT when you use it to create code in your report. Decide on your methods:  use "variable analysis" or other appropriate descriptors.  Make sure to choose at least one graphical method and at least one numerical method.!-->

# Results

<!--Divide this section into two sub-sections:  One for your descriptive  results and one for your inferential results.!-->

## Descriptive Results
```{r}
basicCN <- ggplot(GSScsv,aes(x=DEGREE,y=WORDSUM))
basicCN + geom_boxplot()
```

There is definitely a significant difference displayed within the boxplot as GRADUATE’s have a significantly higher first quartile than the HIGH SCHOOL’s third quartile. The same can be said with the JUNIOR COLLEGE’s first quartile is slightly underneath the LT HIGH SCHOOL’s third quartile

### Graphical Descriptive Results

<!--Graphical results here.  Make sure to show your code.  Provide appropriate labels for axes, giving units if possible, and provide a good title for the graph, too.  Use the graphical results to describe the patterns if any that exist in the data as focused toward the research question!-->

### Numerical Descriptive Results

```{r}
favstats(WORDSUM~DEGREE,data=GSScsv)
```    

GRADUATE has a 01 of 10 compared to the Q3 of LT HIGH SCHOOL is 8.25 and the Q9 of High school is & despite its high Numbers of 85 people. This is important since the number of people in HIGH SCHOOL is almost the combined number of all other DEGREE $ showing that there is great difference between these values.   

<!--Numerical results go here. Use the numerical results to describe the patterns if any that exist in the data as focused toward the research question!-->

## Inferential Results

```{r}
model1 <- aov(WORDSUM~DEGREE,data=GSScsv)
summary.aov(model1)
```

The P-Value is 0.000236 meaning there is a statistical diffrence within these results so further testing is required to see the diffrence

```{r}
#we specify the margins starting at the bottom and proceeding clockwise - thus 8.7 is the left side margin to hold the names of the factors
par(mar = c(3, 17, 2, 2))
plot(TukeyHSD(model1, conf.level = 0.98), las=2,)
```

The difference is obvious within this graph as the LT HIGH SCHOOL-GRADUATE and the HIGH SCHOOL GRADUATE are below 0 showing that there is a significant difference when it is compared to the other DEGREEs.


<!--State hypothesis clearly.  Make sure your discussion of the inferential test covers all the aspects that the test output produces, such as test statistic, p-value etc.  Make a decision about the null hypothesis, explain the assumptions on which the selected test/procedure was based, and why the chosen procedure satisfys the assumptions and is appropriate to answer the research question!-->


# Discussion and Conclusion

A relationship was found within this data, only found when we visually saw a difference with the GRADUATE’s being significantly higher than the HIGH SCHOOL DEGREE and the IT HIGH SCHOOL DEGREE. To fully see this we saw the numbers showing that there was a difference as the 01 of the GRADUATE was 10 while the 03 of LT HIGH SCHOOL is 8.25 and the 03 of HIGH SCHOOL is 9. This was proven with the one way anova showing a p-value of 3.000236 showing something was happening. After some fiddling around with the code we were able to see a significant difference between two levels within this data those levels being the HIGH SCHOOL and LT HIGH SCHOOL as they are below 0.


<!--Discussion and conclusion here.  If you found a relationship be sure to consider whether the relationship occurs because one of the variavbles causes the other, or whether they perhasps are related for some other reason.  Watch the chapter 6 videos from the GeorgeTown videos collection.!-->
